### 2022-09-05 latex/Chapters/Puzzles.tex
update

%-----------------------------------
%	SUBSECTION 1.1
%-----------------------------------
\subsection{History - The 15 Puzzle}

The first puzzle I will focus on is the sliding puzzle (see \cite{SlidingPuzzleWiki}).The 15-puzzle seems to have been invented in the late 19th century by Noyes Chapman (see \cite{SlidingPuzzleWolfram}), who applied in 1880 for a patent on what was then called "Block Solitaire Puzzle". In 1879, a couple of interesting notes (\cite{Johnson1879}), published in the American Journal of Mathematics proved that exactly half of the $16!$ possible ways of setting up the 16 tiles on the board lead to solvable puzzles. A more modern proof can be found in \cite{Archer1999}.
\\
\\
Since then, several variations of the 15-puzzle have become popular, such as the 24-puzzle. A rather contrived but interesting one is the coiled 15-puzzle (\cite{Coiled15Puzzle}), where the bottom-right and the top-left compartments are adjascent; the additional move that this allows renders all $16!$ configurations solvable.

@@

\subsection{History}
The second puzzle I will focus on is the well known Rubik's Cube (\textbf{RC}), invented in 1974 by Erno Rubik, originally known under the name of \textit{Magic Cube} (see \cite{RubiksWiki}). Since it was commericalised in 1980, it has kown a worldwide success: countless people are playing with the famous original 3x3x3 cube, as well as with countless variations, more or less difficult of it. Competitions nowadays bring together \textit{speedcubers} who have trained for years and memorized various algorithms to solve the Rubik's in astonishly effective and fast times (literally in seconds for the best ones). Interestingly, some of the principles used by speedcubers generalise to different dimensions. As an example, it is quite impressive to watch \textit{Cubastic} solve a 15x15x15 RC in two and a half hours (\cite{151515Rubiks}).

\subsection{Search Space \& Solvability}
In my code as well as in this write-up, as is very customary in the RC literature, I will refer to the faces as F (front), B (back), U (up), D (down), L (left) and R (right). The RC's tiles are of 6 different colors, and without loss of generality, we can consider that these colors are also called F, B, U, D, L and R, and consider the goal state as one where, up to rotations of the whole cube, the color's name match the faces (color F on face F, etc). The RC has, as one would expect, an extremely large state space. If we ignore the equivalence via full-cube rotations, there are 24 goal states, since any of the colors can be placed on the F face, and then any of the 4 adjacent colors can be placed on the U face (say). Once these 2 are chosen, the remaining faces are determined. This is not immediately obvious, but not difficult to convince oneself that this is the case with the following two observations about the structure of corner \textit{cubies}: first observation is that they have 3 determined colors, which are fixed once and for all. In particular, the fact that there is no corner with colors F and B, means that once the RC is in solved state, we have to have colors F and B on opposite faces, and similarly for colors U and D as well as colors L and R. Hence, having fixed the color on face F, the color on face B is fixed too. The second observation is that when facing a corner cubie, the order of its three color (e.g. enumerated clock-wise) is invariant. For instance, consider the corner cubie with colors (F, R, U). There is no scrambling of the cube that will produce clock-wise order of e.g. (F, U, R). This is obvious once you consider that there is no move of the Rubik's cube you could not perform while holding a given corner fixed in space (e.g. by pinching that corner cubie with your fingers and never letting go of it).
\\
Finally a further word of notation. It is common in RC jargon to refer to moves  by the name of the face that is rotated clock-wise (when facing it), and by adding a prime \textit{'} for counter-clock-wise rotations. Sometimes people use a number after the face to indicate repeated similar moves, though I will count in my code that as 2 moves, since obviously e.g. U2 = U U = U' U'. As another example: F B' F2 means rotating the front face clock-wise, the back face counter-clock-wise, followed by the front face twice.

\subsubsection{2 x 2 x 2}

@@

\subsubsection{3 x 3 x 3}

TBD

\subsection{Optimal Cost \& God's Number}

>>>>

%-----------------------------------
%	SUBSECTION 1.1
%-----------------------------------
\subsection{History - the 15-Puzzle}

The first puzzle I will focus on in this thesis is the sliding puzzle (see \cite{SlidingPuzzleWiki}) whose most common variant is the 15-puzzle, seemingly invented in the late 19th century by Noyes Chapman (see \cite{SlidingPuzzleWolfram}), who applied in 1880 for a patent on what was then called "Block Solitaire Puzzle". In 1879, a couple of interesting notes (\cite{Johnson1879}), published in the American Journal of Mathematics proved that exactly half of the $16!$ possible ways of setting up the 16 tiles on the board lead to solvable puzzles. A more modern proof can be found in \cite{Archer1999}.
\\
\\
Since then, several variations of the 15-puzzle have become popular, such as the 24-puzzle. A rather contrived but interesting one is the coiled 15-puzzle (\cite{Coiled15Puzzle}), where the bottom-right and the top-left compartments are adjascent; the additional move that this allows renders all $16!$ configurations solvable.

@@

\subsection{History -- from Erno Rubik to speed cubing competitions}
The second puzzle I will focus on is the well known Rubik's Cube (\textbf{RC}), invented in 1974 by Erno Rubik, originally known under the name of \textit{Magic Cube} (see \cite{RubiksWiki}). Since it was commercialized in 1980, it has kown a worldwide success: countless people are playing with the famous original 3x3x3 cube, as well as with many variants of it, more or less difficult and contrived. Competitions nowadays bring together \textit{speedcubers} who have trained for years and memorized various algorithms to solve the Rubik's in astonishly effective and fast times (literally in seconds for the best ones). Interestingly, some of the principles used by speedcubers generalise to different dimensions. As an example, it is quite impressive to watch \textit{Cubastic} solve a 15x15x15 cube in (only) two and a half hours! (see \cite{151515Rubiks}). The key insight to solving high dimensional \textbf{RC}s (short obviously of coming up with specific and sophisticated strategies) is the following: if we manage to group all the inside (i.e. not located on a corner or an edge of the cube) cubies by colour, and then all the non-corner edges on a given tranche of a face by colour, we are essentially reducing the cube to a 3x3x3, since no face rotation will ever undo the inside cubies and non-corner-edge cubies' grouping. For instance, the following picture shows faces of my own 4x4x4 and 3x3x3 cubes which I have arranged to be in identical configurations.
\\
\\
It is quite interesting to note that the difficuly of hand-solving cubes does not grow nearly as much with increasing dimensionality (most of the time difference comes rather from the difficulty of handling and manipulating the cubes of high dimensions at speed).

\begin{figure}[H]
\centering
\includegraphics[scale=0.3]{./Figures/RC333444}
\decoRule
\caption[Reducing RC]{Reducing a 4x4x4 \textbf{RC} to a 3x3x3 -- home made picture}
\end{figure}

 It is quite clear from that picture above that if we can bring the 4x4x4 cube to such a configuration on all faces (which is not super hard), and we know how to solve a 3x3x3, we are essentially done. (well, almost, since the additional degrees of freedom coming from the higher dimension of the 4x4x4 mean that certain arrangements of the colors which are possible on the 4x4x4 are not on the 3x3x3 (for instance we can never have the center red and white on opposite faces of a 3x3x3, but we easily can on a 4x4x4! Fortunately that is nothing that a couple of easily learnable algorithms of a few moves cannot fix).

\subsection{Conventions: faces, moves and configurations}
In my code as well as in this write-up, I follow the conventions and notations of the \textbf{RC} literature and community as much as possible, I will refer to the faces as F (front), B (back), U (up), D (down), L (left) and R (right). The cubes' tiles are of 6 different colors, and without loss of generality (though the interested reader is referred to appendix \ref{sec:Kociemba} for some subtlety around that), we can consider that these colors are also called F, B, U, D, L and R, and consider the goal state as one where, up to rotations of the whole cube, the color's name match the faces (color F on face F, etc). The Rubik's has, as one would expect, an extremely large state space. I will compute the state space size for the 3x3x3 and 2x2x2 cubes in the next two subsections and discuss conditions under which  a cube is solvable.
\\
It is common in \textbf{RC} jargon to refer to moves  by the name of the face that is rotated clock-wise (when facing it), and by adding a prime \textit{'} for counter-clock-wise rotations. Sometimes people use a number after the face to indicate repeated similar moves, though I will count in my code that as 2 moves, since obviously e.g. U2 = U U = U' U'. As another example: F B' F2 means rotating the front face clock-wise, the back face counter-clock-wise, followed by the front face twice.
\\
Finally, let me make a remark on what I consider in this project to be distinct configurations of the Rubik's cube. Since I am interested in experimenting with, among other methods, \textbf{DRL} and \textbf{DQL}, and that they are usually presented as methods which learn \textit{without-human-knowledge}, I thought it would be a bit akin to cheating to consider that two cubes equivalent by full space rotation are one and the same and represent them as such in my code. Ideally, we would rather let the algorithms learn (if they are able to) about symmetries. For instance, the two following cubes (which are clearly just a rotation away or equivalently a change-in-observer away from one another) are the same for us humans. I however represent them as different tensors in my code, and this is up to the solvers to make sense of things.

\begin{figure}[H]
\centering
\includegraphics[scale=0.4]{./Figures/equivalentcubes}
\caption[Reducing RC]{Two equivalent configurations by full-cube-rotation}
\label{fig:configurations}
\end{figure}

\subsubsection{3x3x3 Rubik's -- search space, solvability \& God's number}

Firstly, it is important to realise that each configuration can be arranged in (only) 24 different ways via full cube rotation in space. In particular that means that there are 24 possible goal states (and not 6! as we might naively think at first), so for instance in figure \ref{fig:configurations} I could only have picked 2 out of 24 different configurations. Why is this?
\\
\\
Any of the six colors can be placed on the F face, and then any of the four adjacent colors can be placed on the U face (say). Once these two colors are chosen, the remaining faces are determined. This is not immediately obvious, but not difficult to convince oneself that this is the case with the following two observations about the structure of corner \textit{cubies}: first observation is that they have 3 determined colors, which are fixed once and for all. In particular, the fact that there is no corner with colors F and B, means that once the cube is in solved state, we have to have colors F and B on opposite faces, and similarly for colors U and D as well as colors L and R. Hence, having fixed the color on face F, the color on face B is fixed too. The second observation is that when facing a corner cubie, the order of its three color (e.g. enumerated clock-wise) is invariant. For instance, consider the corner cubie with colors (F, R, U). No scrambling of the cube will ever produce clock-wise order of e.g. (F, U, R). This is obvious once you consider that there is no move of the Rubik's cube you could not perform while holding a given corner fixed in space (e.g. by pinching that corner cubie with your fingers and never letting go of it).
\\
\\

\subsubsection{2x2x2 Rubik's -- search space, solvability \& God's number}

@@
